% Build: latexmk -pdf main.tex

\documentclass[12pt,a4paper]{article}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[french,english]{babel}
\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage{microtype}
\usepackage{geometry}
\usepackage{hyperref}
\usepackage{mathptmx}

\geometry{margin=2.5cm}

\input{macros}

\title{Stage MIDS}
\author{}
\date{\today}

\begin{document}
\maketitle

\section{Notations}

\begin{itemize}
    \item $X^w_t$ = total number of occurrences of the word $w$ at time $t$
    \item $N_t$ = total number of words in the corpus at time $t$
    \item $f_t = \frac{X^w_t}{N_t}$ = frequency of the word $w$ at time $t$
\end{itemize}



The problem with $f_t$ is that it's not an integer; I would prefer keeping the integer structure of $X_t$, but we still need to use some kind of scaling with respect to $N_t$. For that, I propose to scale the mean of any probabilistic model with respect to $N_t$. 

\section{Spike detection}

\subsection{Step 1: fitting a negative binomial model}

The negative binomial distribution with parameters $p \in (0,1), r >0$ is defined as 
$$\mathbb{P}(X = k) = \frac{\Gamma(k + r)}{\Gamma(k + 1) \Gamma(r)} p^k (1-p)^r. $$
The mean and variance are given by
$$
\mathbb{E}[X] = \frac{r(1-p)}{p} \quad \text{and} \quad \text{Var}[X] = \frac{r(1-p)}{p^2}.
$$
A common parametrization is to set $m=r(1-p)/p$, so that $\mathbb{E}[X] = m$ and $\mathrm{Var}(X) = m + m^2/r$. That's the parametrization we will use, i.e. 
$$\mathbb{P}(X = k) = \frac{\Gamma(k + r)}{\Gamma(k + 1) \Gamma(r)} \left(\frac{r}{m+r}\right)^k \left(\frac{m}{m+r}\right)^r.$$

\begin{itemize}
    \item GPT tells me that fitting a negbin with this parametrization can be done using \texttt{statsmodels.discrete.discrete model.NegativeBinomial}, with $\phi=r^{-1}$.
    \item Tails are exponentiel ie $p_n \approx e^{-c n}$ for some $c$. Tails become ligher when $r$ increases
    \item When $r \to \infty$, the distribution converges to a Poisson distribution with parameter $m$, so this model is really versatile, absorbing Poisson, geometric and Pascal distributions. 
    \item Since $X^w_t$ might heavily depend on the number of words $N_t$, it is natural to scale the model linearly with $N_t$, i.e. parametrizing $X_t \sim \mathrm{NB}(\mu N_t, r)$. It is apparently possible to fit this model for observations $(X_t, N_t)$ using the \texttt{exposure} parameter of \texttt{statsmodels.discrete.discrete model.NegativeBinomial}. In this case $\mu$ will be the mean frequency of the word $w$. 
    \item Once the model is fitted and the optimal $\mu_*, r_*$ are found, it's important to plot the density of $\mathrm{NB}(\mu_* \times 100000, r_*)$ on top of the histogram of the data (the right plot sur les graphes d'Elias) to see if it fits or no
\end{itemize}




\bigskip 

Diagnostics for the quality of the fitting? I don't know. Maybe check the mean and variance and skewness and kurtosis ?

\subsection{Step 2: detecting spikes}

One we fit a model, say $\mu_*,r_*$, we can detect outliers in the data. Note $P_t$ the probability density of $\mathrm{NB}(\mu_* N_t,r_*)$. We set 
$$p_t = P_t([X_t, +\infty]), $$
which is the $p$-value of $X_t$ under the fitted model. We declare $X_t$ an outlier if $p$ is too small, say $p_t < 0.0001$. 

Remarks. 
\begin{itemize}
\item $\log_{10}(p_t)$ is called the "surprise" of the observation $X_t$. The higher, the more surprising.
\item we can use multiple-rounds: fit a first NB2 model, then remove all the "blatant outliers" (the ones with a super high suprise), then refit a model on the remaining data, then remove the outliers again, etc. 
\item Diag: plot the histogram of $p_t$, should roughly be uniform except for a few outliers too close to 0?
\end{itemize}

\subsection{Step 3: clusters of abnormal activity}

\subsection{Step 4: time series ?}

If the above doesnt work very well we could add an autoregressive model on the parameters. @Elias could be nice to plot the autocorrelation function of $X_t$ or rather $f_t$; I strongly expect that there will be some interesting autocorrelations in $f_t$.

Typically we would fit something like $m_t = N_t \times e^{R_t}$ with $R_t$ an AR(1) process?

\end{document}
